---
title: "traitmatch"
author: "I. Bartomeus"
date: "1 Aug 2015"
output: html_document
---
# traitmatch: Package to predict trait matching from species interactions.

This document reproduces the analysis done in Bartomeus et al 2015 (Functional Ecology) to show how it works.

To install the package run (only once):

```{r}
install.packages("devtools") 
require(devtools)
install_github("traitmatch", "ibartomeus")
require(traitmatch)
```

We first use data on predator and prey body size from Barnes et al. 2008. We load the data and logtransform it. The original data is multiplied by 10 to avoid problems with the integral of the normal distribution

```{r}
# Load data for Pred/Prey
fish <- read.table("data/Barnes2008.txt", header = TRUE)

# Each row represents an interaction
head(fish) 

# Define the vectors for the predator and prey size
MPred = log10(fish$standardised_predator_length*10)
MPrey = log10(fish$si_prey_length*10)
```

Then we fit the `integrated_model` that integrates neutral and niche constrains. We use pairwise Prey-Predatory interactions (Tlevel vectors) and we asume that the distribution of preys is well characterized by this network (we use mean and standard deviation directly from the distribution of prey interactions). Here we constrain the parameters with a priori information. For example, the slope of the relationship has to be positive (the bigger the predator, the bigger the prey). Tuning the a priori assumptions is important to get good estimates of the parameters. With unconstrained estimates (MLE), some model optimization will struggle to find maximum likelihood estimates and yield erronous parameter estimates, such as a vertical slope. It is therefore critical to check that the results make sense before accepting them as the MLE. This process is slow, specially for large datasets like this one. You can use the max.time (in seconds) to cut the process to e.g. 900 secoonds (15 minuts). The default is 30 minutes. Note that the estimates can be bad if too little time allowed for the simulated annealing to converge. 

```{r}
?fit_it
pars_pre <- fit_it(integrated_model, 
                   Tlevel1 = MPrey,  
                   Tlevel2 = MPred,
                   mean_Tlevel1 = mean(MPrey),
                   sd_Tlevel1 = sd(MPrey),
                   pars = c(a0 = 0, a1 = 0, b0 = 0, b1 = 0),
                   par_lo = c(a0 = -10, a1 = 0, b0 = -10, b1 = -10),
                   par_hi = c(a0 = 10, a1 = 10, b0 = 10, b1 = 10),
                   max.time = 900)
pars_pre 
```

We can plot the predicted model:

```{r}
# With this model, the first estimate is the intercept and second number the slope.

?plot_pred
plot_pred(pars = pars_pre, Tlevel1 = MPrey, 
          Tlevel2 = MPred, xlab = "log (Predator body size)", 
          ylab = "log (prey body size)", pch = ".")
```

You may want to compare that with a pure niche model (not taking into account the abundances). We can fit the niche models as follow:

```{r}
pars_pre_niche <- fit_it(niche_model,
                         Tlevel1 = MPrey, 
                         Tlevel2 = MPred,
                         mean_Tlevel1 = mean(MPrey),
                         sd_Tlevel1 = sd(MPrey),
                         pars = c(a0 = 0, a1 = 0, b0 = 0, b1 = 0),
                         par_lo = c(a0 = -10, a1 = 0, b0 = -10, b1 = -10),
                         par_hi = c(a0 = 10, a1 = 10, b0 = 10, b1 = 10))

plot_pred(pars = pars_pre_niche, Tlevel1 = MPrey, 
          Tlevel2 = MPred, xlab = "log (Predator body size)", ylab = "log (prey body size)",
           pch = ".")
```

We can see in this case that the estimates are similar, but that the slope (alpha1) is flatter and the range a bit wider. We can compare the accuracy of the integrated, niche and neutral models by comparing their likelihood. 

```{r}
# Compute likelihoods
lh_model <- -integrated_model(pars_pre, MPrey, MPred, mean(MPrey),sd(MPrey))
lh_niche <- -niche_model(pars_pre_niche, MPrey, MPred, mean(MPrey), sd(MPrey))
lh_neutral <- -neutral_model(MPrey, MPred, mean(MPrey), sd(MPrey))

# Visualization
barplot(c(lh_model, lh_niche, lh_neutral), names.arg = c("integrated", "niche", "neutral"))
l1 <- c("Predation", lh_model, lh_niche, lh_neutral)
```

In this case, the integrated model is slightly better than the niche model, and the neutral model the worst.

Now we do the same for grasshopper hervibore data.

```{r}
# Load data for grasshoppers
grass = read.table("data/Deraison2014.txt", h = TRUE)
head(grass)

# As is experimental data, we subset only the "presences" for illustrative purposes.
grass <- subset(grass, Herbivory > 0)

# We first fit a binary model, interaction/no interaction.
pars_grass_bin <- fit_it(integrated_model,
                         Tlevel1 = grass$P.Leaf.dry.matter.content, 
                         Tlevel2 = grass$G.Incisive.strength,
                         mean_Tlevel1 = mean(grass$P.Leaf.dry.matter.content),
                         sd_Tlevel1 = sd(grass$P.Leaf.dry.matter.content)) 

# Note that in this case we use the default pars, which are quite unconstrained, but work well in this case.
pars_grass_bin

plot_pred(pars = pars_grass_bin, Tlevel1 = jitter(grass$P.Leaf.dry.matter.content), 
          Tlevel2 = jitter(grass$G.Incisive.strength), xlab = "Incisive strength", 
          ylab = "Leaf dry matter content")

# Compare with the likelihood of the three models
pars_grass_bin_niche = fit_it(niche_model, 
                   Tlevel1 = grass$P.Leaf.dry.matter.content, 
                   Tlevel2 = grass$G.Incisive.strength,
                   mean_Tlevel1 = mean(grass$P.Leaf.dry.matter.content),
                   sd_Tlevel1 = sd(grass$P.Leaf.dry.matter.content))

# Plot_pred(pars = pars_grass_bin_niche, Tlevel1 = jitter(grass$P.Leaf.dry.matter.content), 
#         Tlevel2 = jitter(grass$G.Incisive.strength), xlab = "Incisive strength", ylab = "Leaf dry matter content")

# Likelihoods
lh_model <- -integrated_model(pars_grass_bin, grass$P.Leaf.dry.matter.content, 
                              grass$G.Incisive.strength, 
                              mean(grass$P.Leaf.dry.matter.content), 
                              sd(grass$P.Leaf.dry.matter.content))
lh_niche <- -niche_model(pars_grass_bin_niche, grass$P.Leaf.dry.matter.content, 
                         grass$G.Incisive.strength, 
                         mean(grass$P.Leaf.dry.matter.content), 
                         sd(grass$P.Leaf.dry.matter.content))
lh_neutral <- -neutral_model(grass$P.Leaf.dry.matter.content, grass$G.Incisive.strength, 
                             mean(grass$P.Leaf.dry.matter.content), 
                             sd(grass$P.Leaf.dry.matter.content))

barplot(c(lh_model, lh_niche, lh_neutral), names.arg = c("integrated", "niche", "neutral"))
l2 <- c("Hervibory_bin", lh_model, lh_niche, lh_neutral)

# Here the integrated is as good as the neutral model.

# Now we test the same data taking into account the frequency of interactions.

# Prepare the data
head(grass)
Incisive.strength <- c()
for(i in 1:nrow(grass)){
  temp <- rep(grass$G.Incisive.strength[i], round(grass$Herbivory[i]))
  Incisive.strength <- c(Incisive.strength,temp)
}
Dry.matter <- c()
for(i in 1:nrow(grass)){
  temp <- rep(grass$P.Leaf.dry.matter.content[i], round(grass$Herbivory[i]))
  Dry.matter <- c(Dry.matter,temp)
}

# Fit models
pars_grass_freq <- fit_it(integrated_model,
                         Tlevel1 = Dry.matter, 
                         Tlevel2 = Incisive.strength,
                         mean_Tlevel1 = mean(grass$P.Leaf.dry.matter.content),
                         sd_Tlevel1 = sd(grass$P.Leaf.dry.matter.content))

# Note the distribution (mean and standard deviation) is unweigthed because the experiment had equal abundances of plant species. We can use this knowledge to atribute each plant equal weight.
pars_grass_freq
plot_pred(pars = pars_grass_freq, Tlevel1 = jitter(Dry.matter,100), 
          Tlevel2 = jitter(Incisive.strength), xlab = "Incisive strength", ylab = "Leaf dry matter content")

# Compare models
pars_grass_freq_niche <- fit_it(niche_model, 
                                Tlevel1 = Dry.matter, 
                                Tlevel2 = Incisive.strength,
                                mean_Tlevel1 = mean(grass$P.Leaf.dry.matter.content),
                                sd_Tlevel1 = sd(grass$P.Leaf.dry.matter.content))

# plot_pred(pars = pars_grass_freq_niche, Tlevel1 = jitter(Dry.matter,100), 
 #         Tlevel2 = jitter(Incisive.strength), xlab = "Incisive strength", ylab = "Leaf dry matter content")

# Likelihood
lh_model <- -integrated_model(pars_grass_freq, Dry.matter, Incisive.strength, 
                              mean(grass$P.Leaf.dry.matter.content), 
                              sd(grass$P.Leaf.dry.matter.content))
lh_niche <- -niche_model(pars_grass_freq_niche, Dry.matter, Incisive.strength, 
                         mean(grass$P.Leaf.dry.matter.content), 
                         sd(grass$P.Leaf.dry.matter.content))
lh_neutral <- -neutral_model(Dry.matter, Incisive.strength, 
                             mean(grass$P.Leaf.dry.matter.content),
                             sd(grass$P.Leaf.dry.matter.content))

# Visualization
barplot(c(lh_model, lh_niche, lh_neutral), names.arg = c("integrated", "niche", "neutral"))
l3 <- c("Hervibory_freq", lh_model, lh_niche, lh_neutral)
```

And here is the code for the plants and pollinators:

```{r}
# Read data
pols = read.table("data/Bartomeus2008.txt", h = TRUE)
head(pols)

# Transform body size to tongue lenght first based on Cariveau et al. Submitted.
require(devtools)
install_github("BeeIT", "ibartomeus")
library(BeeIT)
head(pols)
pols$tongue <- ITtongue(pols$IT_mm,pols$family , mouthpart = "tongue")

#plot(pols$tongue, pols$IT_mm) #see the relationship if interested
#subset the plant data for later use.
plants <- unique(pols[,c("plant", "nectar_holder_depth_mm", "cover")])

# Fit models
pars_pols <- fit_it(integrated_model,
                    Tlevel1 = pols$nectar_holder_depth_mm, 
                      Tlevel2 = log(pols$tongue),
                    mean_Tlevel1 = weighted_mean(plants$nectar_holder_depth_mm, plants$cover),
                    sd_Tlevel1 = weighted_sd(plants$nectar_holder_depth_mm, plants$cover))

# Note that here we have independent data on plant abundance (% cover). Hence we can use this data directly to fit the model.
# If interested, you can see that the inference from the network, is quite similar to the real cover 
#plot(dnorm(x = seq(-7,10,1), weighted_mean(plants$nectar_holder_depth_mm, plants$cover), 
 #     weighted_sd(plants$nectar_holder_depth_mm, plants$cover)), col = "red", type = "l")
#par(new = TRUE)
#plot(dnorm(x = seq(-7,10,1), mean(pols$nectar_holder_depth_mm), 
 #     sd(pols$nectar_holder_depth_mm)), col = "blue", type = "l")

plot_pred(pars = pars_pols, Tlevel1 = jitter(pols$nectar_holder_depth_mm, 10), 
          Tlevel2 = log(pols$tongue), xlab = "log(Pollinator tongue size)", ylab = "Nectar depth")

# Model comparision
pars_pol_niche = fit_it(niche_model, 
                   Tlevel1 = pols$nectar_holder_depth_mm, 
                   Tlevel2 = log(pols$tongue),
                   mean_Tlevel1 = weighted_mean(plants$nectar_holder_depth_mm, plants$cover),
                   sd_Tlevel1 = weighted_sd(plants$nectar_holder_depth_mm, plants$cover))
plot_pred(pars = pars_pol_niche, Tlevel1 = jitter(pols$nectar_holder_depth_mm, 10), 
          Tlevel2 = log(pols$tongue), xlab = "log(Pollinator tongue size)", ylab = "Nectar depth")

# Likelihood
lh_model <- -integrated_model(pars_pols, pols$nectar_holder_depth_mm, log(pols$tongue), 
                              weighted_mean(plants$nectar_holder_depth_mm, plants$cover), 
                              weighted_sd(plants$nectar_holder_depth_mm, plants$cover))
lh_niche <- -niche_model(pars_pol_niche, pols$nectar_holder_depth_mm, log(pols$tongue), 
                         weighted_mean(plants$nectar_holder_depth_mm, plants$cover),
                         sd(pols$nectar_holder_depth_mm))
lh_neutral <- -neutral_model(pols$nectar_holder_depth_mm, log(pols$tongue), 
                             weighted_mean(plants$nectar_holder_depth_mm, plants$cover),
                             weighted_sd(plants$nectar_holder_depth_mm, plants$cover))

barplot(c(lh_model, lh_niche, lh_neutral), names.arg = c("integrated", "niche", "neutral"))
l5 <- c("Pollintion", lh_model, lh_niche, lh_neutral)
```

Hosts and parasitoids. An example that doesn't work as well.

```{r}
# Read and format data
host = read.table("data/Tylianakis2008.txt", h = TRUE)
head(host)
host_body_length <- c()
for(i in 1:nrow(host)){
  temp <- rep(host$host_body_length[i], host$freq[i])
  host_body_length <- c(host_body_length, temp)
}
parasite_body_length <- c()
for(i in 1:nrow(host)){
  temp <- rep(host$parasite_body_length[i], host$freq[i])
  parasite_body_length <- c(parasite_body_length,temp)
}

# Fit models
pars_host <- fit_it(integrated_model,
                        Tlevel1 = host_body_length, 
                        Tlevel2 = parasite_body_length,
                        mean_Tlevel1 = weighted_mean(host$host_body_length, host$freq),
                        sd_Tlevel1 = weighted_sd(host$host_body_length, host$freq),
                        par_lo = c(a0 = 0, a1 = -10, b0 = -10, b1 = -10),
                        par_hi = c(a0 = 10, a1 = 0, b0 = 10, b1 = 10))

# With unconstrained parameters, the model behaves wierd, so we have to constrain the values to sensible limits. You can do that by looking at x and y axes ranges and think which slope values are possible.
pars_host
plot_pred(pars = pars_host, Tlevel1 = jitter(host_body_length), 
          Tlevel2 = jitter(parasite_body_length), xlab = "Parasite body size", ylab = "Host body size")
# As you can see everything is predicted.

# Model comparison 
pars_host_niche <- fit_it(niche_model,
                    Tlevel1 = host_body_length, 
                    Tlevel2 = parasite_body_length,
                    mean_Tlevel1 = weighted_mean(host$host_body_length, host$freq),
                    sd_Tlevel1 = weighted_sd(host$host_body_length, host$freq),
                    par_lo = c(a0 = 0, a1 = -10, b0 = -10, b1 = -10),
                    par_hi = c(a0 = 10, a1 = 0, b0 = 10, b1 = 10))

plot_pred(pars = pars_host_niche, Tlevel1 = host_body_length, 
          Tlevel2 = parasite_body_length, xlab = "Parasite body size", ylab = "Host body size")

# Likelihoods
lh_model <- -integrated_model(pars_host, host_body_length, parasite_body_length, 
                              weighted_mean(host$host_body_length, host$freq), 
                              weighted_sd(host$host_body_length, host$freq))
lh_niche <- -niche_model(pars_host_niche, host_body_length, parasite_body_length, 
                         weighted_mean(host$host_body_length, host$freq), 
                         weighted_sd(host$host_body_length, host$freq))
lh_neutral <- -neutral_model(host_body_length, parasite_body_length, 
                             weighted_mean(host$host_body_length, host$freq), 
                             weighted_sd(host$host_body_length, host$freq))

barplot(c(lh_model, lh_niche, lh_neutral), names.arg = c("integrated", "niche", "neutral"))
l4 <- c("Parasitsim", lh_model, lh_niche, lh_neutral)

```
Thats's it. You can also gather a summary table of parameters and likelihoods for all models done:

```{r}
# Table of likelihoods
d <- rbind(l1,l2,l3,l4,l5)
d <- as.data.frame(d)
colnames(d) <- c("system", "integrated", "niche", "neutral")
d

dd <- rbind(pars_pre, pars_pre_niche, pars_grass_bin, pars_grass_bin_niche, pars_grass_freq, 
            pars_grass_freq_niche, pars_host, pars_host_niche, pars_pols, pars_pol_niche)
dd <- as.data.frame(dd)
colnames(dd) <- c("a0", "a1", "b0", "b1")
dd

```

